### 2022-04-11 chapters/fairness/justicia/framework.tex
revised upto practical setting

\noindent\makebox[\linewidth]{\rule{\paperwidth}{2.5pt}}

We use the \red{probability of positive prediction}s of the most and least favored groups to compute different fairness metrics. We next prove the equivalence of {\justiciaenum} and {\justicialearn} in Lemma~\ref{fairness_justicia_lm:equivalence}.

\begin{lemma}
	\label{fairness_justicia_lm:equivalence}
	Let $ \Phi_{\mathbf{a}} $ be the RE-SSAT formula for computing the \red{probability of positive prediction} of the compound sensitive group $ \mathbf{a} \in A $. If $ \Phi_{\mathsf{ER}} $ is the ER-SSAT formula for learning the most favored group and $ \Phi_{\mathsf{UR}} $ is the UR-SSAT formula for learning the least favored group, then
	$\max_{\mathbf{a}} \; \Pr[\Phi_{\mathbf{a}}] = \Pr[\Phi_{\mathsf{ER}}]$   
	and
	$\min_{\mathbf{a}} \; \Pr[\Phi_{\mathbf{a}}] = \Pr[\Phi_{\mathsf{UR}}]$.   
\end{lemma}
\iffalse
\begin{theorem}
	For an ER-SSAT problem, the na\"ive sample complexity is given by 
	\[ k = O\left(\frac{1}{\epsilon^2} (n + \ln(1/\delta))  \right)\]
	where $\hat{p} - p \leq \epsilon$ with probability $1-\delta$.
\end{theorem}
\red{For a simple sampling based algo after evaluating $k$ assignments, the error between estimates probability and actual probability would differ by $\epsilon$ with probability $1- 2^{n+1} e^{-2\epsilon^2 k}$. So the sample complexity of a naive algo is $k = O(\epsilon^{-2} (\log(1/\delta) + n))$.}
\fi

@@

\subsection{Theoretical Analysis: Error Bounds}\label{fairness_justicia_sec:theory}
We access the data generating distribution through  finite number of samples observed from it. These finite sample set introduce errors in the computed probabilities of the randomised quantifiers being $1$. These finite-sample errors in computed probabilities induce further errors in the computed positive predictive value (\red{probability of positive prediction}) and fairness metrics. We next provide a bound on this finite-sample error.

Let us consider that $\hat{p_i}$ is the estimated probability of a Boolean variable $ \bool_i $ being assigned to $ 1 $ from $k$-samples and $p_i$ is the true probability according to $ \mathcal{D} $. 
%If $ | p_i - \hat{p_i}| \le \epsilon $, i.e., $ \epsilon $ additive error for small $ \epsilon \approx 0 $, we want to compute the error of probability $ p $ of the satisfaction of the SSAT formula $ \Phi $.  
Thus, the true satisfying probability $p$ of $ \Phi $ is the weighted sum of all satisfying assignments of the CNF $ \phi $: $p = \sum_\sigma \prod_{\bool_i \in \sigma}p_i$.
This probability is estimated as $\hat{p}$ using $k$-samples from the data generating distribution $\mathcal{D}$ such that $\hat{p} \leq \epsilon_0 p$ for $\epsilon_0 \geq 1$. 
%However, the true satisfying probability $ \hat{p} $  of $ \Phi $ is 
%\[
%\hat{p} = \sum_\sigma \prod_{\bool_i \in \sigma}\hat{p_i}  \leq \sum_\sigma \prod_{\bool_i \in \sigma}(1 \pm \epsilon_i) p_i = \prod_{i =1}^m (1 \pm \epsilon_i) p = \epsilon_0 p
%\]
\iffalse
If we consider $\epsilon_i = \frac{\ln \epsilon_0}{2^i}$, we obtain
\begin{equation*}
\begin{split}
\prod_{i =1}^m (1 \pm \epsilon_i) &\leq \left(\frac{1}{n} \sum_{i} (1 + \epsilon_i)\right)^n \\
&= (\frac{1}{n} \sum_{i} (1+\frac{\ln \epsilon_0}{2^i})^n\\
&\leq (1  +\frac{\ln \epsilon_0}{n})^n\\
&\leq e^{\ln \epsilon_0} = \epsilon_0.\notag
\end{split}
\end{equation*}
\fi
\begin{theorem}\label{fairness_justicia_thm:sample}
	For an ER-SSAT problem, the sample complexity is given by 
	$ k = O\left((n+ \ln(1/\delta))\frac{\ln m}{\ln \epsilon_0} \right),$
	where $\frac{\hat{p}}{p} \leq \epsilon_0$ with probability $1-\delta$ such that $\epsilon_0 \geq 1$.
\end{theorem}
%Theorem~\ref{fairness_justicia_thm:sample} states that for the prescribed number of samples the estimated disparate impact $\hat{DI}$ and statistical parity $\hat{SP}$ would also satisfy $\hat{DI} \leq  \epsilon_0 DI, $ and $\hat{SP} \leq 2\epsilon_0 SP$.
%Apply Hoeffding inequality for each term and then use Union bound for $2^n$ existential variables.
\begin{corollary}
	If $k$ samples are considered from the data-generating distribution in {\justicia} such that 
	$
	k = O\left((n+ \ln(1/\delta))\frac{\ln m}{\ln \epsilon_0}\right),
	$
	the estimated disparate impact $\hat{DI}$ and statistical parity $\hat{SP}$ satisfy, with probability $1-\delta$,
	$
	\hat{DI} \leq  \epsilon_0 DI, \quad \text{and} \quad \hat{SP} \leq 2\epsilon_0 SP.
	$
\end{corollary}
This implies that given a classifier $ \alg \triangleq \Pr(\hat{Y}|X, A) $ represented as a CNF formula and a data-generating distribution $ (X,A,Y) \sim \mathcal{D} $, {\justicia} can verify independence and separation notion of fairness up to an error level $\epsilon_0$ and $2\epsilon_0$ with probability $1-\delta$.
Thus, {\justicia} is a sound framework of fairness verification with high probability.
%\red{Proof of all theorems for arxiv version}
%Use definition of $GF = p_1/p_2$.

>>>>

We next prove the equivalence of {\justiciaenum} and {\justicialearn} in Lemma~\ref{fairness_justicia_lm:equivalence}.

\begin{lemma}
	\label{fairness_justicia_lm:equivalence}
	Let $ \Phi_{\mathbf{a}} $ be an RE-SSAT formula for computing the probability of positive prediction of a classifier corresponding to the sensitive group $  \sensitive = \mathbf{a} $. Additionally, for the same classifer, let $ \Phi_{\mathsf{ER}} $ be an ER-SSAT formula for inferring the most favored sensitive group and $ \Phi_{\mathsf{UR}} $ be a UR-SSAT formula for inferring the least favored sensitive group.  Therefore,	$\max_{\mathbf{a} \in \sensitive} \; \Pr[\Phi_{\mathbf{a}}] = \Pr[\Phi_{\mathsf{ER}}]$   	and	$\min_{\mathbf{a} \in \sensitive} \; \Pr[\Phi_{\mathbf{a}}] = \Pr[\Phi_{\mathsf{UR}}]$.\improvement{Rewrite this lemma in Appendix.}   
\end{lemma}
\iffalse
\begin{theorem}
	For an ER-SSAT problem, the na\"ive sample complexity is given by 
	\[ k = O\left(\frac{1}{\epsilon^2} (n + \ln(1/\delta))  \right)\]
	where $\hat{p} - p \leq \epsilon$ with probability $1-\delta$.
\end{theorem}
\red{For a simple sampling based algo after evaluating $k$ assignments, the error between estimates probability and actual probability would differ by $\epsilon$ with probability $1- 2^{n+1} e^{-2\epsilon^2 k}$. So the sample complexity of a naive algo is $k = O(\epsilon^{-2} (\log(1/\delta) + n))$.}
\fi

@@

\subsection{Theoretical Analysis: Error Bounds}\label{fairness_justicia_sec:theory}
For practical fairness problems, we do not have access to the distribution of features rather to a finite-sample dataset. Since the input of {\justicia} is the distribution of features, we compute the distribution from the dataset. The finite-sample dataset introduces errors in the computed probabilities of randomized quantified features being assigned to $1$. These finite-sample errors in computed probabilities induce further errors in the computed probability of positive prediction of the classifier and corresponding fairness metrics. We next provide a bound on this finite-sample error.

Let us consider that $\hat{p_i}$ is the estimated probability of a Boolean variable $ \bool_i $ being assigned to $ 1 $ from $k$-samples and $p_i$ is the true probability according to $ \mathcal{D} $. 
%If $ | p_i - \hat{p_i}| \le \epsilon $, i.e., $ \epsilon $ additive error for small $ \epsilon \approx 0 $, we want to compute the error of probability $ p $ of the satisfaction of the SSAT formula $ \Phi $.  
\red{Thus, the true satisfying probability $p$ of an SSAT formula $ \Phi $ is the weighted sum of all satisfying assignments of the CNF $ \phi $: $p = \sum_\sigma \prod_{l_i \in \sigma}l_i * p_i + (1 - l_i) * (1 - p_i)$, where the literal $ l_i \in \{0,1\} $ denotes either the variable $ B_i $ (when $ l_i = 1 $) or its negation (when $ l_i = 0 $). }This probability $ p $ is estimated as $\hat{p}$ using $k$-samples from the distribution $\mathcal{D}$ such that $\hat{p} \leq \epsilon_0 p$ for $\epsilon_0 \geq 1$. 
%However, the true satisfying probability $ \hat{p} $  of $ \Phi $ is 
%\[
%\hat{p} = \sum_\sigma \prod_{\bool_i \in \sigma}\hat{p_i}  \leq \sum_\sigma \prod_{\bool_i \in \sigma}(1 \pm \epsilon_i) p_i = \prod_{i =1}^m (1 \pm \epsilon_i) p = \epsilon_0 p
%\]
\iffalse
If we consider $\epsilon_i = \frac{\ln \epsilon_0}{2^i}$, we obtain
\begin{equation*}
\begin{split}
\prod_{i =1}^m (1 \pm \epsilon_i) &\leq \left(\frac{1}{n} \sum_{i} (1 + \epsilon_i)\right)^n \\
&= (\frac{1}{n} \sum_{i} (1+\frac{\ln \epsilon_0}{2^i})^n\\
&\leq (1  +\frac{\ln \epsilon_0}{n})^n\\
&\leq e^{\ln \epsilon_0} = \epsilon_0.\notag
\end{split}
\end{equation*}
\fi
\begin{theorem}\label{fairness_justicia_thm:sample}
	For an ER-SSAT problem, the sample complexity is given by 
	$ k = O\left((n+ \ln(1/\delta))\frac{\ln m}{\ln \epsilon_0} \right),$
	where $\frac{\hat{p}}{p} \leq \epsilon_0$ with probability $1-\delta$ such that $\epsilon_0 \geq 1$.
\end{theorem}
%Theorem~\ref{fairness_justicia_thm:sample} states that for the prescribed number of samples the estimated disparate impact $\hat{DI}$ and statistical parity $\hat{SP}$ would also satisfy $\hat{DI} \leq  \epsilon_0 DI, $ and $\hat{SP} \leq 2\epsilon_0 SP$.
%Apply Hoeffding inequality for each term and then use Union bound for $2^n$ existential variables.
\begin{corollary}
	If $k$ samples are considered from the  distribution in {\justicia} such that 
	$
	k = O\left((n+ \ln(1/\delta))\frac{\ln m}{\ln \epsilon_0}\right),
	$
	the estimated disparate impact $\hat{DI}$ and statistical parity $\hat{SP}$ satisfy, with probability $1-\delta$,
	$
	\hat{DI} \leq  \epsilon_0 DI, \quad \text{and} \quad \hat{SP} \leq 2\epsilon_0 SP.
	$
\end{corollary}
This implies that given a binary classifier represented as a CNF formula and a distribution of sensitive and non-sensitive features, {\justicia} can verify independence and separation notion of fairness up to an error level $\epsilon_0$ and $2\epsilon_0$ with probability $1-\delta$.
Thus, {\justicia} is a sound framework of fairness verification with high probability.
%\red{Proof of all theorems for arxiv version}
%Use definition of $GF = p_1/p_2$.

### 2022-04-11 chapters/fairness/justicia/practical.tex
revised upto practical setting

\subsection{Practical Settings}
\label{sec:practical-setting}
We now relax the assumptions on access to Boolean classifiers and Boolean attributes, and extend {\justicia} to verify fairness metrics for more practical settings of decision trees, linear classifiers, and continuous attributes.

\subsubsection{Extending to Decision Trees and Linear Classifiers. }
In the SSAT approach, we assume that the classifier $\alg$ is represented as a CNF formula.  
We extend {\justicia} beyond CNF classifiers to decision trees and linear classifiers, which are widely used in the fairness studies~\cite{zemel2013learning,raff2018fair,zhang2019faht}.
%{\color{red}In the literature of interpretable machine learning, several studies have been conducted for learning CNF classifiers in the supervised learning setting, which include but are not limited to the work of~\cite{angelino2017learning,malioutov2018mlic,ghosh19incremental}.  We leverage these techniques. (not needed)}

\textit{Binary decision trees} are trivially encoded as  CNF formulas.  In the binary decision tree, each node in the tree is a literal. A \textit{path from the root to the leaf} is a conjunction of literals and thus, a \textit{clause}. The \textit{tree} itself is a disjunction of all paths and thus, a \textit{DNF (Disjunctive Normal Form)}. In order to derive a CNF of a decision tree, we first construct a DNF by including all paths terminating at leaves with negative class label ($ \hat{Y} = 0 $) and then complement the DNF to CNF using De Morgan's rule.

\textit{Linear classifiers on Boolean attributes} are encoded into CNF formulas using pseudo-Boolean encoding~\cite{philipp2015pblib}. We consider a linear classifier  $ W^T X + b \ge 0 $ on Boolean attributes $ X $ with weights $ W \in \mathbb{R}^{|X|} $ and bias $ b \in \mathbb{R} $.  We first normalize $W$ and $b $ in $ [-1,1] $ and then round to integers so that the decision boundary becomes a pseudo-Boolean constraint.  We then apply  pseudo-Boolean constraints to CNF translation to encode the decision boundary to CNF. This encoding usually introduces additional Boolean variables and results in large CNF. In order to generate a smaller CNF, we can trivially apply thresholding  on the weights to consider attributes with higher weights only. For instance, if the weight $  |w_i| \le \lambda $ for a threshold $ \lambda  \in \mathbb{R}^+$ and $ w_i \in W $, we can set $ w_i = 0 $. Thus, the attributes with lower weights and thus, less importance do not appear in the encoded CNF.  Moreover, all introduced variables in this CNF translation are given existential ($ \exists $) quantification and they appear in the inner-most position in the prefix of the SSAT formula. Thus, the presented ER-SSAT formulas become effectively ERE-SSAT formulas.

\subsubsection{Extending to Continuous Attributes.}
In practical problems, attributes are generally real-valued or categorical but classifiers, which are naturally expressed as CNF such as~\cite{GMM20}, are generally trained on a Boolean abstraction of the input attributes.
In order to perform this Boolean abstraction, each categorical attribute is one-hot encoded and each real-valued attribute is discretised into a set of Boolean attributes~\cite{LKCL2019,GMM20}.

For a binary decision tree, each attribute, including the continuous ones, is compared against a constant at each internal node of the tree. We fix a Boolean variable for each internal node, where the Boolean assignment to the variable decides one of the two branches to choose from the current node.

Linear classifiers are generally trained on continuous attributes, where we apply the following discretization. 
Let us consider a continuous attribute $X_c$, where $w$ is its weight during training. 
We discretize $ X_c $ to a set $ \mathbf{B} $ of Boolean attributes and recalculate the weight of each variable in $ \mathbf{B} $ based on $ w $. 
For the discretization of $X_c$, we consider the interval-based approach\footnote{Our implementation is agnostic to any discretization technique.}. 
For each interval in the continuous space of $X_c$, we consider a Boolean variable $B_i \in \mathbf{B}$, such that $ B_i $ is assigned TRUE when the attribute-value of $X_c$ lies within the $i^{\mathrm{th}}$ interval and $ B_i $ is assigned FALSE otherwise. 
Following that, we assign the weight of $ B_i $ to be $ \mu_i\times w $, when $ \mu_i $ is the mean of the $i^{\mathrm{th}}$ interval and  $ B_i $ is TRUE. 
We can show that if we consider infinite number of intervals, $ X_c \approx \sum_i \mu_i B_i $.

>>>>

\subsection{Practical Settings}
\label{sec:practical-setting}
We now relax the assumptions of {\justicia} on access to Boolean classifiers and Boolean features, and extend {\justicia} to verify fairness metrics for more practical settings of decision trees, linear classifiers, and continuous features.

\subsubsection{Extending {\justicia} to Decision Trees and Linear Classifiers. }
In the SSAT approach, we assume that the classifier $\alg$ is represented as a CNF formula.  
We extend {\justicia} beyond CNF classifiers to decision trees and linear classifiers, which are widely used in the fairness studies~\cite{zemel2013learning,raff2018fair,zhang2019faht}.
%{\color{red}In the literature of interpretable machine learning, several studies have been conducted for learning CNF classifiers in the supervised learning setting, which include but are not limited to the work of~\cite{angelino2017learning,malioutov2018mlic,ghosh19incremental}.  We leverage these techniques. (not needed)}

\textit{Binary decision trees} are trivially encoded as  CNF formulas.  In the binary decision tree, each node in the tree is considered as a literal. A \textit{path from the root to the leaf} is a conjunction of literals and thus, a \textit{clause}. The \textit{tree} itself is a disjunction of all paths and thus, a \textit{DNF (Disjunctive Normal Form)}. In order to derive a CNF of a decision tree, we first construct a DNF by including all paths terminating at leaves with negative class label ($ \hat{Y} = 0 $) and then complement the DNF to CNF using De Morgan's rule.

\textit{Linear classifiers on Boolean features} are encoded into CNF formulas using pseudo-Boolean encoding~\cite{philipp2015pblib}. We consider a linear classifier  $ \mathbf{W}^T \mathbf{X} + b \ge 0 $ on Boolean features $ \mathbf{X} $ with weights $ \mathbf{W} \in \mathbb{R}^{|\mathbf{X}|} $ and bias $ b \in \mathbb{R} $.  We first normalize $ \mathbf{W} $ and $ b $ in $ [-1,1] $ and then round to integers so that the decision boundary becomes a pseudo-Boolean constraint~\cite{roussel2009pseudo}.  We then apply  pseudo-Boolean constraints to CNF translation~\cite{philipp2015pblib} to encode the decision boundary to CNF. This encoding usually introduces additional Boolean variables and results in large CNF. In order to generate a smaller CNF, we can trivially apply thresholding  on the weights to consider features with higher weights only. For instance, if the weight $  |w_i| \le \lambda $ for a threshold $ \lambda  \in \mathbb{R}^+$ and $ w_i \in \mathbf{W} $, we can set $ w_i = 0 $. Thus, features with lower weights (less important) do not appear in the encoded CNF.  Moreover, all introduced variables in this CNF translation are given existential ($ \exists $) quantification and they appear in the inner-most position in the prefix of the SSAT formula. Thus, the presented ER-SSAT formulas become effectively ERE-SSAT formulas.

\subsubsection{Extending to Continuous Features.}
In practical problems, features are generally real-valued or categorical but classifiers, which are naturally expressed as CNF such as~\cite{GMM20}, are generally trained on a Boolean abstraction of input features. In order to perform the Boolean abstraction, each categorical feature is one-hot encoded and each real-valued feature is discretized into a set of Boolean features~\cite{LKCL2019,GMM20}.

For a binary decision tree, each feature, including the continuous ones, is compared against a constant at each node (except leaves) of the tree. We assign a Boolean variable to each node (except leaves), where the $ \{0,1\} $ assignment to the variable decides one of the two branches to choose from the current node.

Linear classifiers are generally trained on continuous features, where we apply discretization in the following way. Let us consider a continuous feature $X_c$, where $w$ is its weight during training. We discretize $ X_c $ to a set $ \mathbf{B} $ of Boolean features and recalculate the weight of each variable in $ \mathbf{B} $ based on $ w $. For the discretization of $X_c$, we consider the interval-based approach\footnote{Our implementation is agnostic to any discretization technique.}. For each interval in the continuous space of $X_c$, we consider a Boolean variable $B_i \in \mathbf{B}$, such that $ B_i $ is assigned $ 1 $ when the feature-value of $X_c$ lies within the $i^{\mathrm{th}}$ interval and $ B_i $ is assigned $ 0 $ otherwise. Following that, we assign the weight of $ B_i $ to be $ \mu_iw $, where $ \mu_i $ is the mean of feature values in the $i^{\mathrm{th}}$ interval. We can show that if we consider infinite number of intervals, $ X_c \approx \sum_i \mu_i B_i $.

\noindent\makebox[\linewidth]{\rule{\paperwidth}{2pt}}
